---
title: "Cyclistic 2019 Data Cleaning and Manipulation"
author: "Thomas N"
date: "November 20, 2025"
output: html_document
---
For this case study, the historical data collected by Cyclistic on its users will be used to find 
trends within the various user types and how they use the bike sharing service to help the 
marketing department strategize a plan to get non-members to subscribe to an annual membership. 
This document serves as a record documenting the various steps taken to clean the data, 
along with the findings.

# Table of Contents
- [Libraries Used](#libraries-used)
- [Loading in the Data](#loading-in-the-data)
- [Data Cleaning](#data-cleaning)
- [Data Manipulation](#data-manipulation)
  - [Calculating overall user statistics](#calculating-overall-user-statistics)
  - [Organizing by user type](#organizing-by-user-type)
  - [Individual user type ride stats](#individual-user-type-ride-stats)
    - [Casual users’ average ride length throughout the week](#casual-users-average-ride-length-throughout-the-week)
    - [Members' average ride length throughout the week](#members-average-ride-length-throughout-the-week)
    - [Members'ride count throughout the week](#members-ride-count-throughout-the-week)
    - [Casual users' ride count throughout the week](#casual-users-ride-count-throughout-the-week)
- [Demographic Data](#demographic-data)
  - [Age demographic](#age-demogrphic)
  - [Gender demographic](#gender-demogrphic)
- [Exporting to CSV for analysis](#exporting-to-csv-for-analysis)
- [Visualizing The Data](#visualizing-the-data)
  - [Libraries used for data visualization](#libraries-used-for-data-visualization)
  - [Loading in the data](#loading-in-the-data-1)
  - [Visualizing members vs casual users’ yearly average ride length](#visualizing-members-vs-casual-users-yearly-average-ride-length)
  - [Visualizing weekly ride length average: members vs casual users](#visualizing-weekly-ride-length-average-members-vs-casual-users)
  - [Visualizing weekly ride count: members vs casual users](#visualizing-weekly-ride-count-members-vs-casual-users)
  - [Visualizing casual users' age demographic](#visualizing-casual-users-age-demographic)
  - [Visualizing members' age demographic](#visualizing-members-age-demographic)
  - [Visualizing Gender demographic: casual users vs members](#visualizing-gender-demographic-casual-users-vs-members)
- [Findings](#findings)

# Libraries Used

```{r libraries, echo=TRUE,warning=FALSE,results="hide", message=FALSE}

# loading various libraries to use for analysis
library(tidyverse)
library(dplyr)
library(hms) # used to format time

```

```{r  echo=FALSE, results="hide", message=FALSE}
library(knitr)
```

# Loading in the Data

```{r loading-csv-files, echo=TRUE, results="hide", message=FALSE}


# loading in all the csv files
q1 <- read_csv("./data_2019/Divvy_Trips_2019_Q1.csv")
q2 <- read_csv("./data_2019/Divvy_Trips_2019_Q2.csv")
q3 <- read_csv("./data_2019/Divvy_Trips_2019_Q3.csv")
q4 <- read_csv("./data_2019/Divvy_Trips_2019_Q4.csv")
```

# Data Cleaning

After loading the CSV files, the consistency of the data was inspected by checking to see if all the 
column names were consistent across all four CSV files, along with checking if they contained the 
same number of columns and types of data within them. It was discovered that within Q2's data, 
the column names didn't match the others, and to mend that, all the column names were changed to 
match the other CSV files. After doing so, all 4 data sets were then combined into one data frame 
to represent the entire year for further data cleaning and manipulation.


```{r checking-data-consistency, echo=TRUE}


#-------------checking if all the column names are the same and consistent to combine them
print("q1 columns")
print(colnames(q1))
print("------------")
print("q2 columns")
print(colnames(q2))
print("------------")
print("q3 columns")
print(colnames(q3))
print("------------")
print("q4 columns")
print(colnames(q4))
print("------------")


#-------renaming the column names to match the others----
colnames(q2) <- c(
  "trip_id", "start_time", "end_time", "bikeid",
  "tripduration", "from_station_id", "from_station_name", "to_station_id",
  "to_station_name", "usertype", "gender", "birthyear"
)
#-----------------------------
yearly_df <- rbind(q1, q2, q3, q4) # combining them into 1 data set
```

The Data was further cleaned by checking for any missing or inconsistent data, such as missing values 
or incorrect values. For example, the terms used to describe the two different user types were 
“member” and “casual”; thus, the values in the usertype were changed to match, along with any 
invalid values entered for a user’s birth year. 

```{r data-cleaning, echo=TRUE}
#--------counting number of blank values
na_counts <- colSums(is.na(yearly_df))
print(na_counts)

# renaming values in the columns to keep with consistency of terms used
# in outline
yearly_df$usertype[yearly_df$usertype == "Subscriber"] <- "member"
yearly_df$usertype[yearly_df$usertype == "Customer"] <- "casual"

# renaming the invalid numbers to be NA instead to keep with consistency
yearly_df$birthyear[yearly_df$birthyear == "Invalid Number"] <- "NA"
```


As a precautionary step, the start and end times of the individual trips were checked to see if any 
were entered incorrectly, such as the start and end times are being swapped. 
If any were wrong/swapped the start and end time values were swapped to amend the error.

```{r data-cleaning-wrong-dates, echo=TRUE, warning=FALSE}
wrong <- yearly_df %>% filter(
  as.POSIXct(end_time) < as.POSIXct(start_time)
)
View(head(wrong, 100)) #checking if any dates are swapped

temp <- yearly_df %>%
  mutate(
    started_at_temp = if_else(as.POSIXct(end_time) < as.POSIXct(start_time), end_time, start_time),
    ended_at_temp = if_else(as.POSIXct(end_time) < as.POSIXct(start_time), start_time, end_time),
  )

yearly_df$start_time <- temp$started_at_temp
yearly_df$end_time <- temp$ended_at_temp
```


# Data Manipulation

## Calculating overall user statistics

To calculate the overall user statistics, such as average ride length, maximum ride length for all 
user types and the most popular day of the week all user types use the service, various functions 
were used to perform the calculations. For example, to calculate how long each trip went for, 
the `difftime` function was used to subtract the end time and start time of a trip, represented in 
minutes. Along with that was using the weekdays function to calculate what day the trip took place on, 
as the date recorded followed the ISO 8601 format, which only specifies the year, month, date and 
time it took place. The date the trip ended was used to calculate the day the trip took place, 
assuming users started and ended their usage of the service on the same day.

```{r calculating-usertype-stats, echo=TRUE}
ride_length_secs <- difftime(yearly_df$end_time, yearly_df$start_time, units = "secs")
day_of_week <- weekdays(yearly_df$end_time)
yearly_df$ride_length_minutes <- difftime(yearly_df$end_time, yearly_df$start_time, units = "mins") %>% round(digits = 2)
# formatting trip length, requires time in secs
yearly_df$ride_length_formatted <- as_hms(ride_length_secs)
yearly_df$day_of_week <- day_of_week


View(head(yearly_df, 100))

mean_ride_length <- mean(ride_length_secs) %>% as_hms()
max_ride_length <- max(ride_length_secs) %>% as_hms()
day_of_week_ride_count <- table(yearly_df$day_of_week)
print("weekly trip count all usertypes:")
print(day_of_week_ride_count)

# convert to data frame to get the name of the day of the week from contingency table
mode_day_of_week <- which.max(day_of_week_ride_count) %>%
  as.data.frame()
mode_day_of_week <- rownames(mode_day_of_week)[1] # name for day of the week is used as the row identifier rather than a number
day_of_week_ride_count_df <- as.data.frame(day_of_week_ride_count)


sprintf("mean ride length: %s", mean_ride_length) %>% print()
sprintf("max ride length: %s", max_ride_length) %>% print()
print("------------------------")
sprintf("mode day of the week: %s", mode_day_of_week) %>% print()
print("------------------------")
```

## Organizing by user type

The data was split into two separate data frames, one for members and the other for casual users. 
By doing so, when it came to analyzing the data allowed me to see any trends between 
the two different user types on how they each use the bike-sharing service.


```{r data-manipulation-usertype-dataframes, echo=TRUE}
#-----------members average ride length overall-------
members_df <- yearly_df[yearly_df$usertype == "member", ]
casuals_df <- yearly_df[yearly_df$usertype == "casual", ]
```

After splitting the yearly data frame of all user types into two, the overall yearly average ride 
length of each user type was calculated to see how they differ from each other. 
That data was then combined with the previously calculated stats, such as the average ride 
length of all user types combined, the longest trip taken, and the most popular day of the week 
the service was used, into a single data frame that would later be exported to a CSV file for 
further analysis.

```{r data-manipulation-yearly-stats-dataframe, echo=TRUE}
members_average_trip_length_secs <- mean(as.numeric(members_df$ride_length_minutes)) * 60
casuals_average_trip_length_secs <- mean(as.numeric(casuals_df$ride_length_minutes)) * 60
members_average_trip_length <- members_average_trip_length_secs %>%
  as_hms() %>%
  round_hms(digits = 0)

casuals_average_trip_length <- casuals_average_trip_length_secs %>%
  as_hms() %>%
  round_hms(digits = 0)
sprintf("member average ride length: %s", members_average_trip_length) %>% print()
sprintf("casual average ride length: %s", casuals_average_trip_length) %>% print()
trips_stats_summary_df <- data.frame(
  all_usertypes_average_ride_length = mean_ride_length,
  all_usertypes_max_ride_length = max_ride_length,
  members_average_ride_length = members_average_trip_length,
  casuals_average_ride_length = casuals_average_trip_length,
  mode_day_of_week
)
View(trips_stats_summary_df)
```

```{r echo=FALSE}
kable(trips_stats_summary_df, caption = "Yearly Ride Stats of All User Types") # to display data table 
```

<br>

## Individual user type ride stats

For the two different user types (member, casual), their individual ride stats were calculated based 
on their usage/activity throughout a week. Such as the average ride length grouped by the day of 
the week or the number of trips taken using the bike-sharing service for the two different user types.


### Casual users' average ride length throughout the week

The code chunk below filters through the data frame containing all data for casual users’ 
activity using the bike sharing service and calculates their average ride length grouped by 
the day of the week.


```{r data-manipulation-casual-weekly-average-stats, echo=TRUE}
#------------DATA FRAMES WEEKLY AVERAGE USERTYPES: MEMEBERS, CASUALS-------------------
week_day_column <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

#---------casual weekly average data frame------------
casuals_monday_average <- mean(
  casuals_df$ride_length_minutes[casuals_df$day_of_week == "Monday"]
) %>% round(digits = 2)
casuals_tuesday_average <- mean(
  casuals_df$ride_length_minutes[casuals_df$day_of_week == "Tuesday"]
) %>% round(digits = 2)
casuals_wednesday_average <- mean(
  casuals_df$ride_length_minutes[casuals_df$day_of_week == "Wednesday"]
) %>% round(digits = 2)
casuals_thursday_average <- mean(
  casuals_df$ride_length_minutes[casuals_df$day_of_week == "Thursday"]
) %>% round(digits = 2)
casuals_friday_average <- mean(
  casuals_df$ride_length_minutes[casuals_df$day_of_week == "Friday"]
) %>% round(digits = 2)
casuals_saturday_average <- mean(
  casuals_df$ride_length_minutes[casuals_df$day_of_week == "Saturday"]
) %>% round(digits = 2)
casuals_sunday_average <- mean(
  casuals_df$ride_length_minutes[casuals_df$day_of_week == "Sunday"]
) %>% round(digits = 2)
casuals_weekly_average_ride_length <- c(
  casuals_monday_average,
  casuals_tuesday_average,
  casuals_wednesday_average,
  casuals_thursday_average,
  casuals_friday_average,
  casuals_saturday_average,
  casuals_sunday_average
)
casuals_weekly_ride_average_df <- data.frame(
  day_of_week = week_day_column,
  ride_length_mins = casuals_weekly_average_ride_length
)
View(casuals_weekly_ride_average_df)

```

```{r echo=FALSE}
kable(casuals_weekly_ride_average_df, caption = "Casual Users Weekly Ride Length Average") # display data table
```

### Members' average ride length throughout the week

The code chunk below filters through the data frame containing all the data for the member user type’s 
activity using the bike sharing service and calculates their average ride length grouped by the day 
of the week.


```{r data-manipulation-member-weekly-average-stats, echo=TRUE}
#---memebers weekly average data frame------------
members_monday_average <- mean(
  members_df$ride_length_minutes[members_df$day_of_week == "Monday"]
) %>% round(digits = 2)
members_tuesday_average <- mean(
  members_df$ride_length_minutes[members_df$day_of_week == "Tuesday"]
) %>% round(digits = 2)
members_wednesday_average <- mean(
  members_df$ride_length_minutes[members_df$day_of_week == "Wednesday"]
) %>% round(digits = 2)
members_thursday_average <- mean(
  members_df$ride_length_minutes[members_df$day_of_week == "Thursday"]
) %>% round(digits = 2)
members_friday_average <- mean(
  members_df$ride_length_minutes[members_df$day_of_week == "Friday"]
) %>% round(digits = 2)
members_saturday_average <- mean(
  members_df$ride_length_minutes[members_df$day_of_week == "Saturday"]
) %>% round(digits = 2)
members_sunday_average <- mean(
  members_df$ride_length_minutes[members_df$day_of_week == "Sunday"]
) %>% round(digits = 2)
members_weekly_average_ride_length <- c(
  members_monday_average,
  members_tuesday_average,
  members_wednesday_average,
  members_thursday_average,
  members_friday_average,
  members_saturday_average,
  members_sunday_average
)
members_weekly_ride_average_df <- data.frame(
  day_of_week = week_day_column,
  ride_length_mins = members_weekly_average_ride_length
)
View(members_weekly_ride_average_df)

```

```{r echo=FALSE}
kable(members_weekly_ride_average_df, caption = "Members Weekly Ride Length Average")
```


### Combined data frame of casual users and members' weekly ride length average

Both the members’ and casual users’ weekly ride length average data frames were then combined into a 
single data frame to be exported later as a CSV file for further analysis.


```{r echo=TRUE}
members_casuals_weekly_ride_average_df <- data.frame(
  day_of_week = week_day_column,
  members_ride_length_mins = members_weekly_average_ride_length,
  casuals_ride_length_mins = casuals_weekly_average_ride_length
)
View(members_casuals_weekly_ride_average_df)
```

```{r echo=FALSE}
kable(members_casuals_weekly_ride_average_df, caption = "Members vs Casual Users Average Weekly Ride Length")
```

### Members' ride count throughout the week

The code chunks below calculates amount of trips taken by members throughout the week by filtering and 
grouping trips recorded by the day of the week. After that, the number of trips taken on each 
weekday was summed up by counting how many rows of data were grouped under that day.


```{r data-manipulation-member-weekly-ride-count-stats, echo=TRUE}

#------------DATA FRAMES WEEKLY RIDE COUNT USERTYPES: MEMEBERS, CASUALS-------------------
members_monday_ride_count <- members_df %>%
  filter(day_of_week == "Monday") %>%
  nrow()
members_tuesday_ride_count <- members_df %>%
  filter(day_of_week == "Tuesday") %>%
  nrow()
members_wednesday_ride_count <- members_df %>%
  filter(day_of_week == "Wednesday") %>%
  nrow()
members_thursday_ride_count <- members_df %>%
  filter(day_of_week == "Thursday") %>%
  nrow()
members_friday_ride_count <- members_df %>%
  filter(day_of_week == "Friday") %>%
  nrow()
members_saturday_ride_count <- members_df %>%
  filter(day_of_week == "Saturday") %>%
  nrow()
members_sunday_ride_count <- members_df %>%
  filter(day_of_week == "Sunday") %>%
  nrow()
```

### Casual users' ride count throughout the week

The code chunks below calculates amount of trips taken by casual users throughout the week by filtering 
and grouping the trips recorded by the day of the week. After that, the number of trips taken on 
each weekday was summed up by counting how many rows of data were grouped under that day.

```{r data-manipulation-casual-weekly-ride-count-stats, echo=TRUE}
casuals_monday_ride_count <- casuals_df %>%
  filter(day_of_week == "Monday") %>%
  nrow()
casuals_tuesday_ride_count <- casuals_df %>%
  filter(day_of_week == "Tuesday") %>%
  nrow()
casuals_wednesday_ride_count <- casuals_df %>%
  filter(day_of_week == "Wednesday") %>%
  nrow()
casuals_thursday_ride_count <- casuals_df %>%
  filter(day_of_week == "Thursday") %>%
  nrow()
casuals_friday_ride_count <- casuals_df %>%
  filter(day_of_week == "Friday") %>%
  nrow()
casuals_saturday_ride_count <- casuals_df %>%
  filter(day_of_week == "Saturday") %>%
  nrow()
casuals_sunday_ride_count <- casuals_df %>%
  filter(day_of_week == "Sunday") %>%
  nrow()
```

### Combined data frame of casual users and members' weekly ride count

After creating the individual data frames that calculated the total number of trips taken per day of 
the week for each user type, they were then combined into a single data frame, which would later 
be exported as a CSV file for further analysis.

```{r echo=TRUE}
members_weekly_ride_count <- c(
  members_monday_ride_count, members_tuesday_ride_count,
  members_wednesday_ride_count, members_thursday_ride_count,
  members_friday_ride_count, members_saturday_ride_count,
  members_sunday_ride_count
)

casuals_weekly_ride_count <- c(
  casuals_monday_ride_count, casuals_tuesday_ride_count,
  casuals_wednesday_ride_count, casuals_thursday_ride_count,
  casuals_friday_ride_count, casuals_saturday_ride_count,
  casuals_sunday_ride_count
)
members_casuals_weekly_ride_count_df <- data.frame(
  week_day_column,
  members_weekly_ride_count,
  casuals_weekly_ride_count
)
View(members_casuals_weekly_ride_count_df)
```

```{r echo=FALSE}
kable(members_casuals_weekly_ride_count_df, caption = "Members Vs Casual Users Weekly Ride Count")
```

<br>

# Demographic Data

## Age demographic

The data for the year 2019 also contained demographic data, such as the year each user was born, 
for each trip taken. This was done by counting and grouping the number of trips taken and the 
registered birth year of the user for that trip, while filtering out any rows of data where the 
year of birth was absent/missing.


```{r demographic-dataframes-age, echo=TRUE}
#-------------------------------DEMOGRAPHIC DATA FRAME: BIRTH YEAR------------------
casuals_birth_year_df <- casuals_df %>%
  filter(birthyear != "NA") %>%
  count(birthyear)

View(casuals_birth_year_df)

members_birth_year_df <- members_df %>%
  filter(birthyear != "NA") %>%
  count(birthyear)

View(members_birth_year_df)
```

```{r echo=FALSE}
kable(casuals_birth_year_df, caption = "Casual Users' Demographic Data Frame")
kable(members_birth_year_df, caption = "Members' Demographic Data Frame")
```

## Gender demographic

The data from 2019 also housed information about the gender of a user any time the service was used. 
Two data frames were created to store the registered genders by user type while filtering out any 
missing values and counting the number of occurrences for each unique value.


```{r demographic-dataframes-gender, echo=TRUE}
#---------DEMOGRAPHIC DATA FRAME: GENDER------------------
casuals_gender_df <- casuals_df %>%
  filter(gender != "NA") %>%
  count(gender)
View(casuals_gender_df)

members_gender_df <- members_df %>%
  filter(gender != "NA") %>%
  count(gender)
View(members_gender_df)
```


# Exporting to CSV for Analysis

With the several data frames created from manipulating and cleaning up the data, they were then 
exported to CSV files using the following code for analysis in a separate R script.


```{r exporting-csv, eval=FALSE}
source("./analysis_2019/analysis_2019.R") # source code/file containing the dataframes
# export to csv
write_csv(trips_stats_summary_df, "./summary_results_2019/trips_stats_summary.csv")
write_csv(members_casuals_weekly_ride_count_df, "./summary_results_2019/members_casuals_weekly_ride_count.csv")
write_csv(members_casuals_weekly_ride_average_df, "./summary_results_2019/members_casuals_weekly_ride_average.csv")
write_csv(members_birth_year_df, "./summary_results_2019/members_birth_year.csv")
write_csv(casuals_birth_year_df, "./summary_results_2019/casuals_birth_year.csv")
write_csv(casuals_gender_df, "./summary_results_2019/casuals_gender.csv")
write_csv(members_gender_df, "./summary_results_2019/members_gender.csv")
```

<br>
<br>

# Visualizing The Data

## Libraries used for data visualization

```{r vislualization-libraries, echo=TRUE, warning=FALSE, results="hide", message=FALSE}
library(ggplot2)
library(readr)
library(dplyr)
library(lubridate)
```

## Loading in the data

The various CSV files created previously were then loaded into a separate R 
script with the following code to create data visuals.

```{r visualization-loading-data, echo=TRUE, warning=FALSE, results="hide", message=FALSE}
directory <- "./summary_results_2019"
trips_summary_csv <- read_csv(file.path(directory, "trips_stats_summary.csv"))
weekly_ride_average_csv <- read_csv(file.path(directory, "members_casuals_weekly_ride_average.csv"))
weekly_ride_count_csv <- read_csv(file.path(directory, "members_casuals_weekly_ride_count.csv"))
casuals_birth_year_csv <- read_csv(file.path(directory, "casuals_birth_year.csv"))
members_birth_year_csv <- read_csv(file.path(directory, "members_birth_year.csv"))
casuals_gender_csv <- read_csv(file.path(directory, "casuals_gender.csv"))
members_gender_csv <- read_csv(file.path(directory, "members_gender.csv"))
colnames(casuals_birth_year_csv)[2] <- "count"
colnames(members_birth_year_csv)[2] <- "count"
colnames(casuals_gender_csv)[2] <- "count"
colnames(members_gender_csv)[2] <- "count"
```

## Visualizing members vs casual users' yearly average ride length

Their average ride time was converted from HH:MM:SS format to minutes(1 hour 20 minutes = 80 minutes) 
to easily display and compare the difference between the user types when using a bar graph to 
represent the data.

```{r members-vs-casuals-average-ride-length-bar-graph, echo=TRUE}
members_ride_average_min <- (as.difftime(trips_summary_csv$members_average_ride_length, format = "%H:%M:%S, units = minutes") / 60) %>%
  as.numeric()
casuals_ride_average_min <- (as.difftime(trips_summary_csv$casuals_average_ride_length, format = "%H:%M:%S, units = minutes") / 60) %>%
  as.numeric()
all_usertypes_ride_length_df <- data.frame(
  user_type = c("members", "casuals"),
  ride_length = c(members_ride_average_min, casuals_ride_average_min)
)

View(all_usertypes_ride_length_df)
bar_chart_usertypes_average_ride_length <- ggplot(
  data = all_usertypes_ride_length_df
) +
  geom_col(mapping = aes(
    x = user_type, y = ride_length, fill = user_type
  )) +
  labs(
    title = "Average Ride Length 2019",
    x = "User Type",
    y = "Ride Length (Min)",
    fill = "User Type"
  ) +
  scale_y_continuous(
    breaks = seq(0, max(all_usertypes_ride_length_df$ride_length) + 10, by = 10),
    expand = expansion(add = c(0, 5))
  )
```

```{r echo=FALSE}
print(bar_chart_usertypes_average_ride_length) 
```


## Visualizing weekly ride length average: members vs casual users

To visualize the difference between members and casual users’ average ride length throught a week 
the data was pulled from their respective columns in the members_casuals_weekly_ride_average.csv 
and converted to a long format by creating 2 separate data frames for each user type where each 
row will contain the day of the week, the average ride length on that day and the user type to 
create a stacked area graph that will plot the data by mapping the data to their respective user type.


```{r members-vs-casuals-weekly-ride-length-average-area-graph, echo=TRUE}
weekly_ride_average_len <- nrow(weekly_ride_average_csv)
temp <- c("character", length = nrow(weekly_ride_average_csv))
for (i in 1:weekly_ride_average_len) {
  temp[i] <- "casual"
}

day_of_week <- factor(weekly_ride_average_csv$day_of_week,
  levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
)

casuals_weekly_ride_average_df <- data.frame(
  day_of_week = day_of_week,
  ride_length = c(weekly_ride_average_csv$casuals_ride_length_mins),
  user_type = temp
)

temp2 <- c("character", length = nrow(weekly_ride_average_csv))
for (i in 1:weekly_ride_average_len) {
  temp2[i] <- "member"
}
members_weekly_ride_average_df <- data.frame(
  day_of_week = day_of_week,
  ride_length = c(weekly_ride_average_csv$members_ride_length_mins),
  user_type = temp2
)


weekly_ride_average_df <- rbind(casuals_weekly_ride_average_df, members_weekly_ride_average_df)
View(weekly_ride_average_df)
area_graph_average_weekly_ride_length <- ggplot(weekly_ride_average_df, aes(x = day_of_week, y = ride_length, fill = user_type, group = user_type)) +
  scale_y_continuous(
    breaks = seq(0, round(max(weekly_ride_average_df$ride_length), -1) + 10, by = 10)
  ) +
  geom_area(position = "identity", alpha = 0.7) +
  geom_line(linewidth = 1, stat = "identity", color = "#6b6b6b") +
  geom_point(size = 2, color = "#6b6b6b", stat = "identity") +
  coord_fixed(ratio = 0.05) +
  labs(
    x = "Day of the Week",
    y = "Ride Length (Min)",
    title = "Weekly Ride Length 2019"
  )
```

```{r echo=FALSE}
print(area_graph_average_weekly_ride_length)
```


## Visualizing weekly ride count: members vs casual users

To compare how the number of trips throughout a week by day of the week differs between casual 
users and members, data pulled from the “weekly_ride_count.csv” file was converted from long 
format to wide format as 2 separate columns list the number of rides taken by each user type, 
with each row representing a day of the week. By converting it to wide format, 
it allowed me to create a side-by-side bar chart where the number of rides between both user 
types can be compared per day of the week. This was achieved by creating two vectors, 
one for each user type, where their ride count would be ordered from Monday to Friday and 
then placed into two separate data frames for each user type. The two separate data frames 
will then be joined together into a single one to act as a dataset for the side-by-side bar 
chart to use and visualize the data.

```{r members-vs-casuals-weekly-ride-count-bar-graph, echo=TRUE}


usertype_member <- c("character", length = nrow(weekly_ride_average_csv))
for (i in 1:weekly_ride_average_len) {
  usertype_member[i] <- "member"
}

members_ride_count <- vector("numeric", 7)
for (i in 1:7) {
  members_ride_count[i] <- weekly_ride_count_csv$members_weekly_ride_count[
    weekly_ride_count_csv$week_day_column == day_of_week[i]
  ]
}

members_weekly_ride_count_df <- data.frame(
  day_of_week,
  user_type = usertype_member,
  ride_count = members_ride_count
)

usertype_casual <- c("character", length = nrow(weekly_ride_average_csv))
for (i in 1:weekly_ride_average_len) {
  usertype_casual[i] <- "casual"
}
casuals_ride_count <- vector("numeric", 7)
for (i in 1:7) {
  casuals_ride_count[i] <- weekly_ride_count_csv$casuals_weekly_ride_count[
    weekly_ride_count_csv$week_day_column == day_of_week[i]
  ]
}

casuals_weekly_ride_count_df <- data.frame(
  day_of_week,
  user_type = usertype_casual,
  ride_count = casuals_ride_count
)


weekly_ride_count_df <- rbind(members_weekly_ride_count_df, casuals_weekly_ride_count_df)
View(weekly_ride_count_df)

bar_chart_weekly_ride_count <- ggplot(
  data = weekly_ride_count_df
) +
  geom_col(mapping = aes(
    x = day_of_week, y = ride_count, fill = user_type
  ), position = "dodge") +
  labs(
    title = "Weekly Ride Count 2019",
    x = "Day of the Week",
    y = "Ride Count"
  ) +
  scale_y_continuous(breaks = seq(0, max(weekly_ride_count_df$ride_count), by = 50000))
```

```{r echo=FALSE}
print(bar_chart_weekly_ride_count)
```

## Visualizing casual users' age demographic

The casual users’ age demographic was pulled from the “casual_birth_year.csv” file and plotted 
onto a line graph with the series of birthyears placed on the x-axis and the y-axis being the 
number of trips taken where the registered user’s birth year for that trip falls under a 
particular year.


```{r casuals-age-demographic-line-graph, echo=TRUE}
casuals_line_graph_age_demographic <- ggplot(casuals_birth_year_csv, aes(x = birthyear, y = count)) +
  geom_line(linewidth = 1, stat = "identity") +
  coord_fixed(ratio = 0.009) +
  labs(
    title = "Casual Users' Demographic (Age) 2019",
    x = "Birth Year",
    y = "Count"
  ) +
  scale_x_continuous(
    breaks = seq(round(min(casuals_birth_year_csv$birthyear), digits = -2) - 60, round(max(casuals_birth_year_csv$birthyear), digits = -2) + 10, by = 20),
    expand = c(0, 20)
  )

```

```{r echo=FALSE}
print(casuals_line_graph_age_demographic)
```


## Visualizing members' age demographic

The members’ age demographic was pulled from the “members_birth_year.csv” file and plotted onto 
a line graph with the series of birthyears placed on the x-axis and the y-axis being the number 
of trips taken where the registered user’s birth year for that trip falls under a particular year.

```{r members-age-demographic-line-graph, echo=TRUE}
members_line_graph_age_demographic <- ggplot(members_birth_year_csv, aes(x = birthyear, y = count)) +
  geom_line(linewidth = 1, stat = "identity") +
  coord_fixed(ratio = 0.0007) +
  labs(
    title = "Members' Demographic (Age) 2019",
    x = "Birth Year",
    y = "Count"
  ) +
  scale_x_continuous(
    breaks = seq(round(min(members_birth_year_csv$birthyear), digits = -2) - 20, round(max(members_birth_year_csv$birthyear), digits = -2) + 20, by = 10),
    expand = c(0, 10)
  ) +
  scale_y_continuous(
    breaks = seq(0, round(max(members_birth_year_csv$count), digits = -2), by = 20000)
  )
```

```{r echo=FALSE}
print(members_line_graph_age_demographic)
```

## Visualizing Gender demographic: casual users vs members

Data loaded in from the CSV files created from the filtered casual users and members data sets 
for genders: “casuals_gender.csv”, “members_gender.csv”; was used to calculate the gender ratio 
per user type. After doing so, a side-by-side bar chart was used to visualize how they compare 
against each other.


```{r members-casuals-gender-demographic-bar-chart, echo=TRUE}
#------------------------Gender demographic bar charts---------------

casuals_female_count <- casuals_gender_csv$count[1]
casuals_male_count <- casuals_gender_csv$count[2]
casuals_female_percent <- (casuals_female_count / (casuals_male_count + casuals_female_count)) * 100
casuals_male_percent <- (casuals_male_count / (casuals_male_count + casuals_female_count)) * 100
casuals_female_percent <- round(casuals_female_percent, digits = 2)
casuals_male_percent <- round(casuals_male_percent, digits = 2)

members_female_count <- members_gender_csv$count[1]
members_male_count <- members_gender_csv$count[2]
members_female_percent <- (members_female_count / (members_male_count + members_female_count)) * 100
members_male_percent <- (members_male_count / (members_male_count + members_female_count)) * 100
members_female_percent <- round(members_female_percent, digits = 2)
members_male_percent <- round(members_male_percent, digits = 2)


gender_df <- data.frame(
  gender = c(casuals_gender_csv$gender, members_gender_csv$gender),
  percentage = c(casuals_female_percent, casuals_male_percent, members_female_percent, members_male_percent),
  usertype = c("casual", "casual", "member", "member")
)
View(gender_df)

# custom colour for the bars using scale_fill_manual
bar_chart_gender <- ggplot(gender_df, mapping = aes(x = gender, y = percentage, fill = usertype)) +
  geom_col(
    position = "dodge"
  ) +
  scale_fill_manual(values = c("#57bd70", "#577cbd")) +
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    expand = expansion(mult = 0)
  ) +
  labs(
    title = "Gender Demographic: Members vs Casual Users",
    x = "Gender",
    y = "Percentage"
  ) +
  geom_text(aes(label = percentage, group = usertype, vjust = -1), position = position_dodge(width = 0.9))
```

```{r echo=FALSE}
print(bar_chart_gender)
```

# Findings 

Throughout the year of 2019, casual users tend to have a longer average ride length overall 
compared to members, with an average ride length of 57 minutes and 1 second compared to members 
with an average ride length of 14 minutes and 20 seconds. When comparing the two different user 
types’ weekly average ride length, casual users have Monday as the day of the week with the 
highest average ride time of 69.54 minutes. When comparing it to members, Saturday is the day 
of the week with the highest average ride length for them at 17.09 minutes for them. 

When it comes to the number of times the service is being used, members use the bike-sharing 
more compared to casual users throughout the week, with Tuesday having the highest number of 
recorded trips taken. When compared to casual users, casual users tend to use the service more 
on weekends, with Saturday having the highest number of recorder trips taken by them. 

The year 2019 also had data on the user demographic. Whenever a user takes a trip using the 
bike-sharing service, the user’s gender(male or female) and the year they were born would be 
recorded. Regarding the age demographic, a majority of the users(casual users and members) tend 
to be born between the years of 1990 and 2000. The data has also revealed that the gender 
distribution amongst the user types leans towards more male users compared to female users 
(casual users: 38.19% female, 61.81% male; members: 24.93% female, 75.07% male).
